---
title: "Standardised Classification of Effects Based on Their Uncertainty"
author: Thierry Onkelinx
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Standardised Classification of Effects Based on Their Uncertainty}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(ggplot2)
```

## A Simple, Stable and Transparant Classification of Effects

The `effectclass` package provides a functions to classify effects in a stable and transparant manner.
We base the classification on the lower and upper bounds of a confidence (or credible) interval, focussing on the uncertaincy rather than the point estimates.
First, we compare the confidence interval with a reference.
This reference typically refers to the null hypothesis.
An effect with confidence interval above the reference is a significant (important) effect.
Let's call this a `positive effect`.
Likewise, a `negative effect` has a confidence interval below the reference.

```{r base-effect, fig.cap = "Comparing confidence intervals to the reference.", echo = FALSE}
ds <- data.frame(
  effect = factor(1:2, labels = c("positive\neffect", "negative\neffect")
  ),
  lcl = c(0.1, -0.5),
  ucl = c(0.5, -0.1)
)
ggplot(ds, aes(ymin = lcl, ymax = ucl, x = effect)) +
  geom_errorbar() +
  geom_hline(yintercept = 0, linetype = 2) +
  scale_y_continuous(
    breaks = 0, 
    labels = "reference"
  ) +
  theme(
    axis.title = element_blank()
  )
```

But what when the confidence interval contains the reference?
Then the width of the confidence interval plays an important role.
Suppose the confidence interval is wide.
The value of the effect can range from a strong negative effect over no effect to a strong positive effect.
Basically, we have no information on the effect.
On the otherhand, suppose we have a narrow confidence interval.
The effect stil is either positive or negative, but we know that it won't be a strong positive nor strong negative effect.
We could define a lower^[below the reference] and upper^[above the reference] threshold and use them to separate confidence intervals that contain the reference.
If the confidence interval is between both the upper and lower threshold, we can state that the effect does not exide these thresholds.
We set the thresholds at a level where we can consider the effect to be not relevant so we can call this `no effect`.
As soon as the confidence interval contains the reference and at least on of the threshold, we consider it an `unknown effect`.

```{r base-effect2, fig.cap = "Comparing confidence intervals to the reference and thresholds.", echo = FALSE}
ds <- data.frame(
  effect = factor(1:2, labels = c("no\neffect", "unknown\neffect")
  ),
  lcl = c(-0.2, -0.5),
  ucl = c(0.2, 0.5)
)
ggplot(ds, aes(ymin = lcl, ymax = ucl, x = effect)) +
  geom_errorbar() +
  geom_hline(yintercept = c(0, -0.25, 0.25), linetype = c(2, 3, 3)) +
  scale_y_continuous(
    breaks = c(-0.25, 0, 0.25), 
    labels = c("lower\nthreshold", "reference", "upper\nthreshold")
  ) +
  theme(
    axis.title = element_blank()
  )
```

Comparing the confidence interval with a reference and two thresholds, gives us four distinct categories.
The table below lists the rules for each category and suggests a symbol and standardised wording for effects and trends.

| Symbol | Effect          | Trend    | Rules                                  |
| :----: | --------------- | -------- | -------------------------------------- |
| `+`    | positive effect | increase | confidence interval above reference    |
| `-`    | negative effect | decrease | confidence interval below reference    |
| `~`    | no effect       | stable   | confidence interval between thresholds |
| `?`    | unknown effect  | unknown  | confidence interval contains reference and at least one threshold |

Table: Summary of the simple classification into four categories.
