---
title: "Visualising Effects with `plotly`"
author: Thierry Onkelinx
bibliography: references.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualising Effects with `plotly`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(knitr)
opts_chunk$set(collapse = TRUE, comment = "#>")
inbo_colours <- c("#282A72", "#EF972C")
```

## `add_fan()`

The Bank of England visualises uncertainty by using a fan plot [@britton.fisher.ea1998InflationReportProjections].
Instead of displaying a single coverage interval, they recommend to display a bunch of coverage intervals with different levels of transparency.
`add_fan()` creates the intervals based on a Gaussian distribution.
It uses the `y` aesthetic for the mean and the `sd` as the standard error.
The default setting in `add_fan()` adds intervals in 5% increments.

```{r add-fan-default, fig.cap = "Default plot using `add_fan()`."}
library(effectclass)
library(plotly)
set.seed(20191216)
timeserie <- data.frame(
  year = 1990:2019, dx = rnorm(30, sd = 0.2), s = rnorm(30, 1, 0.05)
)
timeserie$index <- exp(cumsum(timeserie$dx))
plot_ly(timeserie, x = ~year, y = ~index) |>
  add_fan(sd = ~s)
```

You can change the number of intervals by setting the `step` argument.

```{r add-fan-interval, fig.cap = "Fan plot with three intervals and line."}
plot_ly(timeserie, x = ~year, y = ~index) |>
  add_fan(sd = ~s, step = 0.3) |>
  add_lines(color = I(inbo_colours[1]))
```

You can change the width of the interval by setting the `max_prob` argument.

```{r add-fan-max-prob, fig.cap = "Fan plot with a maximum probability of 0.5."}
plot_ly(timeserie, x = ~year, y = ~index) |>
  add_fan(sd = ~s, max_prob = 0.5)
```

You can use the `link` argument to apply a link function.
This assumes that the standard error is on the link scale and the y values in on the natural scale.

```{r add-fan-log, fig.cap = "Fan plot with log link. Note the asymmetric intervals due to the link function."}
plot_ly(timeserie, x = ~year, y = ~index) |>
  add_fan(sd = ~s, link = "log")
```

The next example shows how to add multiple fans to a plot.

```{r add-fan-category, fig.cap = "add_fan() with different colours."}
timeseries <- expand.grid(year = 1990:2019, category = c("A", "B"))
timeseries$dx <- rnorm(60, sd = 0.1)
timeseries$index <- rep(c(0, 2), each = 30) + cumsum(timeseries$dx)
timeseries$s <- rnorm(60, rep(c(0.5, 1), each = 30), 0.05)
plot_ly(x = ~year, y = ~index) |>
  add_fan(
    data = timeseries[timeseries$category == "A", ], sd = ~s,
    fillcolor = inbo_colours[1], name = "A"
  ) |>
  add_fan(
    data = timeseries[timeseries$category == "B", ], sd = ~s,
    fillcolor = inbo_colours[2], name = "B"
  )
```

## `add_vert_norm()`

`add_vert_norm()` adds violin style elements to the plot.
The shape is based on a Gaussian distribution with mean defined by the `y` aesthetic and sigma defined by the `sd` aesthetic.
We symmetrically truncate the distribution so the total probability is equal to `prob`, which is set to `0.95` by default.
We display this distribution vertically and mirrored, so the mean becomes the widest part of the element.
The top and bottom of the element are the extreme percentiles matching the `prob` argument.
When `prob = 0.95`, the top and bottom are the 2.5% and 97.5% percentiles.
Hence, the height of the element can be interpreted as a 95% confidence interval.
Due to the truncation and mirroring, the top and bottom of the element become horizontal lines.

```{r add-vert-norm}
plot_ly(timeserie, x = ~year, y = ~index) |>
  add_vert_norm(sd = ~s)
```

Increasing `prob` will increase the height of the element.
As `prob` approaches 1, there is less truncation and the shape is more recognisable as a Gaussian distribution.
However, you lose the interpretation of the height as a confidence interval in case you want a 90% or 95% confidence interval.

```{r add-vert-norm-9999}
plot_ly(timeserie, x = ~year, y = ~index) |>
  add_vert_norm(sd = ~s, prob = 0.9999)
```

If you want a smaller confidence interval, you can set `prob` to a lower value.

```{r add-vert-norm-80}
plot_ly(timeserie, x = ~year, y = ~index) |>
  add_vert_norm(sd = ~s, prob = 0.8)
```

Setting `truncated = FALSE` will remove the rectangle in the centre of the shape.
The top and bottom of the element now becomes a point instead of a horizontal line.

```{r add-vert-norm-80-t}
plot_ly(timeserie, x = ~year, y = ~index) |>
  add_vert_norm(sd = ~s, prob = 0.8, truncated = FALSE)
```

Setting the `step` argument defines steps in probability at which we approximate the Gaussian distribution.
`prob = 0.9` and `step = 0.45` will create a shape based on the 5%, 50% and 95% percentiles.

```{r add-vert-norm-step}
plot_ly(timeserie, x = ~year, y = ~index) |>
  add_vert_norm(sd = ~s, prob = 0.9, step = 0.45)
```

The `delta` argument defines the half width of the element.
The default value is half of the smallest distance between two consecutive `x` values.

```{r add-vert-norm-delta1}
plot_ly(timeserie, x = ~year, y = ~index) |>
  add_vert_norm(sd = ~s, delta = 1)
```

```{r add-vert-norm-delta02}
plot_ly(timeserie, x = ~year, y = ~index) |>
  add_vert_norm(sd = ~s, delta = 0.2)
```

The `link` argument applies a link function to the `sd` aesthetic.

```{r add-vert-norm-log}
plot_ly(timeserie, x = ~year, y = ~index) |>
  add_vert_norm(sd = ~s, link = "log")
```

## Reference

The `reference_shape()` and `reference_text()` functions can be used to add a reference lines and text to the plot.

```{r reference-fan}
plot_ly(timeserie, x = ~year, y = ~index) |>
  add_fan(sd = ~s) |>
  layout(
    shapes = reference_shape(threshold = 1),
    annotations = reference_text(threshold = 1)
  )
```

```{r reference-vert-norm}
plot_ly(timeserie, x = ~year, y = ~index) |>
  add_vert_norm(sd = ~s) |>
  layout(
    shapes = reference_shape(threshold = 1),
    annotations = reference_text(threshold = 1)
  )
```
