---
title: "Visualising Effects"
author: Thierry Onkelinx
bibliography: references.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualising Effects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(knitr)
opts_chunk$set(collapse = TRUE, comment = "#>")
```

The vignette [classification]("classification.html") explains how we can classify effects based on their confidence intervals.
This vignette focusses on visualising effect and their uncertainty.
The packages provides two `ggplot()` layers (`stat_fan()` and `stat_effect()`) and a scale (`scale_effect()`). 

## `stat_fan()`

The Bank of England visualises uncertainty by using a fan plot [@britton.fisher.ea1998InflationReportProjections].
Instead of displaying a single coverage interval, they recommend to display a bunch of coverage intervals with different levels of transparency.
`stat_fan()` creates the intervals based on a Gaussian distribution.
It uses the `y` aesthetic for the mean and the `link_sd` as the standard error.
The default setting in `stat_fan()` displays three sets of confidence intervals.
Setting `fine = TRUE` yields a set of nine confidence intervals.

| opacity | width around the mean | lower | upper |
| ------: | --------------------: | ----: | ----: |
| 90%     | 30 %                  |  35 % |   65% |
| 60%     | 60 %                  |  20 % |   80% |
| 30%     | 90 %                  |   5 % |   95% |

Table: Definition of the three level intervals

| opacity | width around the mean | lower | upper |
| ------: | --------------------: | ----: | ----: |
| 90%     | 10 %                  |  45 % |   55% |
| 80%     | 20 %                  |  40 % |   60% |
| 70%     | 30 %                  |  35 % |   65% |
| 60%     | 40 %                  |  30 % |   70% |
| 50%     | 50 %                  |  25 % |   75% |
| 40%     | 60 %                  |  20 % |   80% |
| 30%     | 70 %                  |  15 % |   85% |
| 20%     | 80 %                  |  10 % |   90% |
| 10%     | 90 %                  |   5 % |   95% |

Table: Definition of the nine level intervals

```{r stat-fan-default, fig.cap = "Default plot using stat_fan()."}
library(effectclass)
library(ggplot2)
set.seed(20191216)
z <- data.frame(year = 1990:2019, dx = rnorm(30, sd = 0.2), s = rnorm(30, 1, 0.05))
z$index <- exp(cumsum(z$dx))
ggplot(z, aes(x = year, y = index, link_sd = s)) + stat_fan()
```

```{r stat-fan-fine, fig.cap = "Fan plot with nine intervals and line."}
ggplot(z, aes(x = year, y = index, link_sd = s)) + 
  stat_fan(fine = TRUE) + geom_line()
```

Some statistical analyses require a different distribution, e.g. Poisson or binomial.
These analyses often use a link function between the linear predictor and the mean of the distribution.
Poisson uses a log link, binomial a logit link.
The uncertainty around the predictions of the linear predictor (in the link scale) still follows a Gaussian distribution.
However, we want to display the predictions and their uncertainty in the original scale.
The `link` argument `stat_fan()` allows the user to specify the required link function.
`stat_fan()` then assumes that `y` specifies the median in the original scale. 
Hence the confidence intervals in the link-scale use $link(y)$ as their mean.
You need to specify the standard error in the link scale, that why we named the argument `link_sd`.
`stat_fan()` calculates the intervals based on $\mathcal{N}(\mu = link(y), \sigma = link\_sd)$ and backtransforms them to the original scale.
Having `y` in the original scale in combination with a link function has the benefit that it is easy to reuse the `y` in other geoms.

```{r stat-fan-log, fig.cap = "Plot using stat_fan() with log-link. Note the asymmetric intervals due to the link function."}
ggplot(z, aes(x = year, y = index, link_sd = s)) + 
  stat_fan(link = "log") + 
  geom_line()
```

`stat_fan()` handles different discrete colours as demontrated in the example below.

```{r stat-fan-colour, fig.cap = "stat_fan() with different colours."}
z <- expand.grid(year = 1990:2019, category = c("A", "B"))
z$dx <- rnorm(60, sd = 0.1)
z$index <- rep(c(0, 2), each = 30) + cumsum(z$dx)
z$s <- rnorm(60, rep(c(0.5, 1), each = 30), 0.05)
ggplot(z, aes(x = year, y = index, link_sd = s)) + 
  stat_fan(aes(fill = category)) + geom_line(aes(colour = category))
```

## References
